Finding Motifs in Knowledge Graphs using Compression

### 2021-04-16 kgmotifs/kgmotifs.tex
...

\documentclass[letterpaper]{article} %DO NOT CHANGE THIS

\usepackage{aaai19}    % Required
\usepackage{times}     % Required
\usepackage{helvet}    % Required
\usepackage{courier}   % Required
\usepackage{url}       % Required
\usepackage{graphicx}  % Required
\frenchspacing         % Required
\setlength{\pdfpagewidth}{8.5in}  % Required
\setlength{\pdfpageheight}{11in}  % Required
\usepackage{makecell}
\usepackage{arydshln}
\usepackage{float}

% PDF Info Is Required:
\pdfinfo{
/Title (Detecting Motifs in Knowledge Graphs using Compression)
/Author (...)}

\setcounter{secnumdepth}{0}

@@

@@

@@

\title{Detecting Motifs in Knowledge Graphs using Compression}
\author{
Author names withheld
% Peter Bloem\\
% Knowledge Representation \& Reasoning Group\\
% VU University\\
% vu@peterbloem.nl\\
}

@@

@@

\begin{abstract}
\noindent We introduce a method to detect \emph{network motifs} in knowledge graphs. Network motifs are useful patterns or meaningful subunits of the graph that recur frequently. We extend the common definition of a network motif to coincide with a \emph{basic graph pattern}, commonly used in the knowledge graph literature. We introduce an approach, inspired by recent work for simple graphs, for inducing these for a given knowledge knowledge graph, and show that the motifs returned, reflect the basic structure of the graph. Specifically we show that in random graphs, no motifs are found, and that when we insert a motif artificially, it can be detected, even if as few as 10 instances are inserted. Finally, we show the results of motif induction on three real-world knowledge graphs, and show that our method provides good suggestions for motifs that reflect the structure of the graph in an intuitive way.
\end{abstract}

\noindent \emph{Knowledge graphs} are an extremely versatile way of storing knowledge. They allow knowledge to be encoded without a predefined schema, they allow different amounts of data to be encoded for different entities and they allow different modalities to be integrated seamlessly \cite{wilcke2017knowledge}. This versatility comes at a price. For a given knowledge graph, it can be difficult to see the forest for the trees: how is the graph structured at the lowest level? What kind of things can I ask of what types of entities? What are small, recurring patterns that might represent a novel insight into the data? Answering these questions could benefit problem domains like graph simplification, graph navigation and schema induction.

@@

\begin{figure*}[bth]
  \centering
    \includegraphics[width=\textwidth]{example.pdf}
    \caption{An example of the principle behind our motif code. a) A basic knowledge graph. Note that in our analysis, we consider only the indices of the nodes and relations, not their labels. b) A motif that occurs frequently. c) A compressed representation; we remove all edges that are part of an occurrence of the motif. We then store separately which nodes match the motif. The motif, together with the template graph and the matches can be used to reconstruct the graph. We take this as a heuristic: the better the motif compresses, the more characteristic it is likely to be for the graph.}
    \label{figure:example}
  \label{figure:codes}
\end{figure*}

@@

In this paper, we extend the compression-based motif analysis to knowledge graphs. For the purposes of this research we define knowledge graphs as labeled, directed multigraphs. Nodes are uniquely labeled with entity names, and links are non-uniquely labeled with relations.  We extend the definition of a motif to that of a \emph{basic graph pattern}: that is, a motif is a small graph with \emph{some} of its nodes and links labeled with entities and relations as they occur in the graph. Other nodes and links are labeled as \emph{variables}. A pattern matches at a particular location in the graph, if the variables can be replaced with values from the graph, and the motif becomes a subgraph as a result. 
The intuition behind our method is that the more compactly a pattern compresses the graph, the better the motif. Figure~\ref{figure:example} shows an illustration of the basic principle. In the preliminaries section below, we justify this intuition more formally.

We perform several experiments to show that our method returns meaningful subgraphs. First we test the intuition that a random graph should contain no motifs. We also show that when we artificially insert motifs into a random graph, we can then detect these as motifs, sometimes with as few as 10 instances inserted into the graph. Finally, we show the results of motif analysis on three real-world knowledge graphs.

All code and datasets used in this paper are available, under open licenses.\footnotemark

\footnotetext{URL withheld for purposes of blind reviewing.}

@@

Network motifs for unlabeled simple graphs were introduced in \cite{milo2002network}. A more comprehensive overview of the related literature can be found in \cite[Section~1.1]{bloem2017large}. In \cite{bloem2017large}, the principle of Minimum Description Length (MDL) was first connected to motif analysis. However, the idea had earlier been exploited for detecting meaningful subgraphs in the SUBDUE algorithm \cite{cook1994substructure}.

A few other methods have been proposed for inducing the structure of a given knowledge graph. In \cite{pham2015deriving}, the authors use the principle of characteristic sets to characterize a knowledge graph in terms of the star patterns it contains. In \cite{pham2016exploiting}, they show that the majority of the LOD cloud can be efficiently described using such principles, showing the highly tabular structure of many knowledge graphs. In \cite{volker2011statistical}, association rules mining is used to induce basic patterns in the graph.

@@

\paragraph{Minimum Description Length}
Our method is based on the MDL principle: that we should favour models that compress the data. We will show briefly how this intuition can be made mathematically precise. For more details, we refer the reader to \cite{grunwald2007minimum} for MDL in general, and to \cite{bloem2017large}, for a more extensive discussion these principles in the domain of graph analysis.

Let $\B$ be the set of all finite-length binary strings. We use $|b|$ to represent the length of $b \in \B$. Let $\log(x) = \log_2(x)$. A \emph{code} for a set of objects $\cal X$ (such as graphs) is an injective function $f: {\cal X} \to \B$, mapping objects to binary code words. All codes in this paper are \emph{prefix-free}: no code word is the prefix of another. We will denote a \emph{codelength function} with the letter $L$, ie. $L(x) = |f(x)|$. We commonly compute $L(x)$ directly, without first computing $f(x)$ and we refer to $L(x)$ as a code.

A well known result in information theory is the association between codes and probability distributions, implied by the \emph{Kraft inequality}: for each probability distribution $p$ on $\cal X$, there exists a prefix-free code $L$ such that for all $x \in \cal X$: $- \log p(x) \leq L(x) < -\log p(x) + 1$. Inversely, for every prefix-free code $L$ for $\cal X$, there exists a probability distribution $p$ such that for all $x \in \cal X$: $p(x) = 2^{-L(x)}$. For proofs, see \cite[Section~3.2.1]{grunwald2007minimum} or \cite[Theorem~5.2.1]{cover2006elements}. To explain the intuition, note that we can easily transform a code $L$ into a sampling algorithm for $p$ by feeding the decoding function random bits until it produces an output. To transform a probability distribution to a code, techniques like arithmetic coding \cite{rissanen1979arithmetic} can be used.\footnotemark 
\footnotetext{
As explained in \cite[page 96]{grunwald2007minimum}, the fact that $-\log p^*(x)$ is real-valued and $L^*(x)$ is integer-valued can be safely ignored and we may \emph{identify} codes with probability distributions, allowing codes to take non-integer values. 
}

@@

\paragraph{Relevance testing} We will use the MDL principle to perform a hypothesis test. Assume we have some data $x$ and a null hypothesis that it was sampled from distribution $p^\text{null}$. Let $L^\text{null}$ be the corresponding codelength. A simple but crucial result, known as the \emph{no-hypercompression inequality} \cite[p103]{grunwald2007minimum} tells us that the probability of sampling data $x$ from $p^\text{null}$ that can be described shorter than $L^\text{null}(x)$ by $k$ or more bits, \emph{using any code} is less than $2^{-k}$.
Thus, an MDL hypothesis test would consist of stating the null hypothesis in terms of a code $L^\text{null}$, designing a code (before seeing the data) which compresses the data better than $L^\text{null}$ by, say, 10 bits and rejecting the null hypothesis with confidence $2^{-10}$. For a longer, more intuitive explanation of this principle in pattern induction, we refer the reader to \cite[Section~2]{bloem2017large}.

Note that when we use this procedure to find motifs, we are not providing evidence for the hypothesis that the motif is ``correct'' \cite[Section~2]{bloem2017large}.
\paragraph{Common codes}
In the construction of our codes for graphs, we require some simpler codes as building blocks. First, when we store any positive integer $n$, we do so with the code corresponding to the distribution $p^\N(n) = 1/(n(n+1))$, and denote it $L^\N(n)$. For nonnegative numbers we add 1 to the argument. for the full range of integers, we add an extra bit for the sign, and then use the first code for negative integers and the second for positive ones.

We will often need to encode sequences of integers. These will be highly skewed, with only a subset of integers occurring frequently, and others occurring infrequently or not at all. As noted in \cite{de2016names} a code based on the Pitman-Yor model \cite{pitman1997two} is very effective in such situations. Let $S = \langle S_1, ..., S_n\rangle$ be a sequence of integers of length $n$. We first store its members $m(s)$ in the order in which they first occur: we first store $n$ and the first member using the codes described above. The rest of the members are stored sequentially using the distance from the current to the next member.

Let $f(A, B)$ be the frequency of symbol $A$ in sequence $B$.We then store the complete sequence using the code corresponding to the following distribution:
\begin{align*}
&p^\text{PY}_{\alpha,d}(S) = \prod_{i \in [1,k]} \text{PY}_{\alpha,d}(S_i\mid S_{1:i-1}) \\
&\text{PY}_{\alpha,d}(S_i \mid S') = 
\begin{cases}
\frac{\alpha - d |m(S)|}{|m(S)| + \alpha} & \text{if} f(S_i, S') = 0 \\ 
\frac{f(S_i, S') - d}{|m(S)| + \alpha} & \text{otherwise}
\end{cases}
\end{align*}
See \cite{de2016names} for a more intuitive explanation. In all experiments we use $\alpha = 0.5$, $d=0.1$. We will refer to the total codelength for a sequence under this procedure as $L^{PY}(S)$

\section{Method}
To comfortably define probabilities and codes on graphs, we will slightly deviate from conventional definitions. Specifically, we will analyse the \emph{structure} of knowledge graphs only, ignoring any meaning they have outside the graph. Under this assumption we can assume that the node and link labels of our knowledge graphs are simply natural numbers.\footnotemark~This leads to the following definition.

\footnotetext{For practitioners this restriction is not noticeable, as the indices can simply be mapped back to the original strings when the found motifs are presented.}

A \emph{knowledge graph} $G$, hereafter simply a \emph{graph}, is a tuple $G = (v_G, r_G, E_G)$. $v_G \in \N$ is the number of nodes in the graph, and $r_G \in \N$ is the number of relations. We define the nodeset as $V_G = \{0, \ldots, v_G-1\}$ and the relation-set as $R_G = \{0, \ldots, r_G\}$. The \emph{tripleset} $E_G \subset V_G \times R \times V_G$ determines the edges of the graph and their labels.

As shown in Figure~\ref{figure:example}, the patterns that we aim to find are partially labeled: some nodes and edges are labeled with as variables. We will represent these with negative integers. A pattern $M$ for graph $G$ is a tuple $(V_M, R_M, G, E_M)$. Let $v_M$ and $r_M$ indicate the number of variable nodes and variable links in $M$ respectively, then $V_M \subseteq \{-v'_M, \ldots, v_G-1\}$ and $R_M \subseteq \{-(r'_M+v'M), \ldots,-v'M, 0,\ldots, r_G-1\}$, with $E_M \subset V_M \times R_M \times V_M$, as for regular graphs. That is; nodes can be labeled either with nonnegative integers referring to $G$'s nodes or with negative integers representing a variable node, and similar for relations. The negative integers are always contiguous within a single graph, with the highest representing the node labels and the lowest representing the edge labels.
\begin{table}[H]x
\centering
\tiny
\begin{tabular}{r r p{0.5\linewidth}}
\hline
log-factor & frequency & \\
\hline
\multicolumn{3}{c}{Dogfood , top 5 by log-factor ($>100$ positive)}\\
\hline	
 83119.1 & 9344 & \makecell{\texttt{?n1 dc:creator ?n2.} \\\texttt{?n2 rdfs:label ?n3.} \\\texttt{?n2 foaf:name ?n3.} } \\ 
 \hdashline
 81075.1 & 10322 & \makecell{\texttt{?n1 foaf:maker ?n2.} \\\texttt{?n2 rdfs:label ?n3.} \\\texttt{?n2 foaf:name ?n3.} } \\
  \hdashline 
 77692.6 & 5234 & \makecell{\texttt{?n2 rdf:type ?n3.} \\\texttt{?n2 rdfs:label ?n1.} \\\texttt{?n2 foaf:name ?n1.} } \\ 
  \hdashline
 75632.3 & 3536 & \makecell{\texttt{?n1 swc:isPartOf ?n2.} \\\texttt{?n2 swc:hasPart ?n1.} } \\ 
  \hdashline
 75592.4 & 3536 & \makecell{\texttt{?n1 swc:isPartOf ?n2.} \\\texttt{?n2 ?p3 ?n1.} } \\ 
\hline
\multicolumn{3}{c}{Dogfood, top 3 by frequency}\\
\hline	
 -3032.5 & 137008 & \makecell{\texttt{?n1 rdf:\_1 ?n2.} \\\texttt{?n1 rdf:\_2 ?n4.} \\\texttt{?n1 rdf:\_3 ?n3.} } \\ 
  \hdashline
 -3251.4 & 109315 & \makecell{\texttt{?n1 swc:isPartOf ?n3.} \\\texttt{?n2 swc:isPartOf ?n3.} } \\ 
 \hdashline
 2089.0 & 92165 & \makecell{\texttt{?n2 swc:isPartOf ?n3.} \\\texttt{?n3 swc:hasPart ?n1.} } \\ 
\hline
\multicolumn{3}{c}{AIFB, top 5 by log-factor ($>100$ positive)}\\
\hline	
 79234.0 & 7728 & \makecell{\texttt{?n1 ?p3 ?n2.} \\\texttt{?n2 ?p4 ?n1.} } \\ \hdashline
 61310.4 & 4154 & \makecell{\texttt{?n1 swrs:publication ?n2.} \\\texttt{?n2 ?p3 ?n1.} } \\ \hdashline
 57641.1 & 3965 & \makecell{\texttt{?n1 swrs:publication ?n2.} \\\texttt{?n2 swrs:author ?n1.} } \\ \hdashline
 57603.1 & 3965 & \makecell{\texttt{?n1 swrs:author ?n2.} \\\texttt{?n2 ?p3 ?n1.} } \\ \hdashline
 42110.1 & 17512 & \makecell{\texttt{?n1 swrs:publication ?n3.} \\\texttt{?n3 swrs:author ?n2.} \\\texttt{?n3 swrs:author ?n1.} \\\texttt{?n2 swrs:publication ?n3.} } \\ 
\hline
\multicolumn{3}{c}{AIFB, top 3 by frequency}\\
\hline
 2287.7 & 198668 & \makecell{\texttt{?n1 swrs:publication ?n3.} \\\texttt{?n1 swrs:publication ?n2.} } \\ \hdashline
 7822.1 & 191083 & \makecell{\texttt{?n1 swrs:author ?n2.} \\\texttt{?n2 swrs:publication ?n3.} } \\ \hdashline
 1264.0 & 190636 & \makecell{\texttt{?n1 swrs:isAbout ?n3.} \\\texttt{?n2 swrs:isAbout ?n3.} } \\ 
\hline
\multicolumn{3}{c}{Mutag, top 5 by log-factor ($87$ positive)}\\
\hline
 178304.4 & 18634 & \makecell{\texttt{?n1 mtg:\_hasAtom ?n3.} \\\texttt{?n1 mtg:\_hasBond ?n2.} \\\texttt{?n2 mtg:\_inBond ?n3.} } \\ \hdashline
 97237.8 & 9189 & \makecell{\texttt{?n1 mtg:\_hasAtom ?n2.} \\\texttt{?n2 mtg:\_charge ?n3.} } \\ \hdashline
 90447.5 & 18634 & \makecell{\texttt{?n1 mtg:\_hasBond ?n2.} \\\texttt{?n2 mtg:\_inBond ?n4.} \\\texttt{?n2 mtg:\_inBond ?n3.} } \\ \hdashline
 79052.0 & 9317 & \makecell{\texttt{?n1 mtg:\_hasBond ?n2.} \\\texttt{?n2 rdf:type ?n3.} } \\ \hdashline
 58973.9 & 7892 & \makecell{\texttt{?n1 mtg:\_hasAtom ?n2.} \\\texttt{?n2 rdf:type ?n3.} } \\ 
\hline
\multicolumn{3}{c}{Mutag, top 3 by frequency}\\
\hline
 1025.7 & 341143 & \makecell{\texttt{?n2 mtg:\_hasAtom ?n3.} \\\texttt{?n2 mtg:\_hasBond ?n1.} } \\ \hdashline
 -374.5 & 178837 & \makecell{\texttt{?n1 mtg:\_charge ?n3.} \\\texttt{?n2 mtg:\_charge ?n3.} } \\ \hdashline
 -2088.0 & 52985 & \makecell{\texttt{?n1 rdf:type owl:Class.} \\\texttt{?n3 rdfs:subClassOf ?n2.} \\\texttt{?n4 ?p5 owl:Class.} \\\texttt{?n4 rdfs:subClassOf ?n2.} } \\ 
\hline
\end{tabular}
\caption{Results of the experiment on real-world data. For each experiment we also report the number of motifs found with a positive log-factor.}
\label{table:topk}
\end{table}
An \emph{instance} for $M$ in $G$ is a pair of sequences of integers: $I = (I^n, I^r)$. $I^n$ is a sequence of distinct integers of length $v'_M$. $I^r$ is a sequence of non-distinct integers of length $r'_M$. For each edge $(s, p, o) \in E_M$ containing a negative $s$, $p$ or $o$, there is a corresponding link in $E_G$ with a negative $s$ replaced by $I^n_{-s}$, a negative $o$ replaced by $I^n_{-o}$, and a negative $p$ replaced by $I^r_{-p - v'_M}$. Put simply: for a pattern to match, variable edges marked with the same negative integer, must map to the same relation in order for the pattern to match, but variable links labeled with different negative integers \emph{may} map to the same relation. Variable nodes are always labeled distinctly and may never map to the same node in $G$. An instance describes a subgraph of $G$ that \emph{matches} the pattern $M$. Each edge in the motif may only match one edge in the graph. In other words, the occurrence of the motif in the graph must have as many edges as the motif itself.

We will first assume that a target pattern $M$ is given for the data $G$ and that we have a set of instances $\I$. Moreover, we require that all instances in $\I$ are mutually disjoint: no two subgraphs defined by member of $\I$ may share an edge, but nodes may be shared.

\paragraph{Null model}
The most common null model for classical motif analysis is the degree-sequence model (also known a the configuration model \cite{newman2010networks}): a uniform distribution over all graphs with a particular degree sequence. We extend this to knowledge graphs by also including the degree of each relation: that is, the frequency with which it occurs in the tripleset. Let a \emph{degree sequence} $D$ of length $n$ be a triple of three integer sequences: $(D^\text{in}, D^\text{rel}, D^\text{out})$. If $D$ is the degree sequence of a graph, then node $i$ has $D^\text{in}_i$ incoming links,  $D^\text{out}_i$ outgoing links and for each relation $r$, there are $D^\text{rel}_r$ triples.

Let $\G_D$ be the set of all graphs with degree sequence $D$. Then the configuration model can be expressed simply as
\[
p^C(G) = \frac{1}{|\G_D|}
\]
for any $G$ that satisfies $D$ and $p(G) = 0$ otherwise. Unfortunately, there is no efficient way to compute $|\G_D|$ and even approximations tend to be costly for large graphs. Following the approach in \cite{bloem2017large}, we define a fast approximation to the configuration model, which works well in practice for motif detection.

@@

We can thus use 
\[
p^\text{EL}_D(G) =  \frac{m!}{{m \choose {D_1^\text{out}, \ldots, D_n^\text{out}} }
 {m \choose {D_1^\text{rel}, \ldots, D_{|R_G|}^\text{rel}} }
 {m \choose {D_1^\text{in}, \ldots, D_n^\text{in}} }} \leq p^C(G) \text{.}
\]

Filling in the definition of the multinomial coefficient, and rewriting, we get a codelength of:
\begin{align*}
- &\log p^\text{EL}_D(G) = 2 \log(m!) \\
&- \sum_i \log(D_i^\text{in}!)- \sum_i \log(D_i^\text{rel}!)- \sum_i \log(D_i^\text{out}!)
\end{align*}

as an approximation for the DS model. We call this the edgelist (EL) model. It gives a probability that always lower-bounds the configuration model, since it affords some probability mass to graphs that cannot exist.\footnotemark ~Experiments in the classical motif setting have shown that the EL model is an acceptable proxy for the DS model \cite{bloem2017large}, especially considering the extra scalability it affords.

\footnotetext{Note that we cannot simply think of $p^\text{EL}$ as a uniform model for graphs containing multiple links, since we can only divide by $m!$ if we know that all triples are unique.}

>>>>

\documentclass[runningheads]{style/llncs}

\usepackage{url}       
\usepackage{graphicx}  
\usepackage{float}

\usepackage{makecell}
\usepackage{arydshln}

@@

\usepackage{booktabs}

@@

\newcommand{\p}{\;\text{.}}

@@

@@

\title{Finding Motifs in Knowledge Graphs using Compression}
\titlerunning{Finding Motifs in Knowledge Graphs}

\author{Peter Bloem\\ \orcidID{0000-0002-0189-5817}}
% 
%
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Knowledge Representation and Reasoning group,\\ Vrije Universiteit Amsterdam \\
\email{vu@peterbloem.nl} \\
\url{peterbloem.nl}}

@@

\begin{abstract}
\noindent We introduce a method to find \emph{network motifs} in knowledge graphs. Network motifs are useful patterns or meaningful subunits of the graph that recur frequently. We extend the common definition of a network motif to coincide with a \emph{basic graph pattern}. We introduce an approach, inspired by recent work for simple graphs, to induce these from a given knowledge knowledge graph, and show that the motifs found reflect the basic structure of the graph. Specifically, we show that in random graphs, no motifs are found, and that when we insert a motif artificially, it can be detected. Finally, we show the results of motif induction on three real-world knowledge graphs.
\end{abstract}

\section{Introduction}
\noindent \emph{Knowledge graphs} are an extremely versatile and flexible data model. They allow knowledge to be encoded without a predefined format and they are extremely robust in the face of missing data. This versatility comes at a price. For a given knowledge graph, it can be difficult to see the forest for the trees: how is the graph structured at the lowest level? What kind of things can I ask of what types of entities? What are small, recurring patterns that might represent a novel insight into the data? Answering these questions could benefit problem domains like graph simplification, graph navigation and schema induction.

@@

\begin{figure*}[bth]
  \centering
    \includegraphics[width=\textwidth]{example.pdf}
    \caption{An example of the principle behind our motif code. a) A basic knowledge graph. We consider only the indices of the nodes and relations, not their labels. b) A motif that occurs frequently. c) A compressed representation; we remove all edges that are part of an occurrence of the motif and store separately which nodes match the motif.}
    \label{figure:example}
  \label{figure:codes}
\end{figure*}

@@

In this paper, we extend this compression-based motif analysis to \emph{knowledge graphs}. For the purposes of this research we define knowledge graphs as labeled, directed multigraphs. Nodes are uniquely labeled with entity names, and links are non-uniquely labeled with relations.  We extend the definition of a motif to that of a \emph{basic graph pattern}: a small graph labeled with both variables and explicit entitites and relations. A pattern matches if the variables can be replaced with specific values from the graph so that the pattern becomes a subgraph as a result.

The intuition behind our method is that we can use graph patterns to \emph{compress the graph}: we store the pattern, its instances, and the remainder of the graph. The better this representation compresses the graph, the more relevant the pattern. Figure~\ref{figure:example} illustrates the principle. In Section~\ref{section:preliminaries}, we justify this intuition more formally.

We perform several experiments to show that our method returns meaningful subgraphs. First we test the intuition that a random graph should contain no motifs. We also show that when we artificially insert motifs into a random graph, we can then detect these as motifs. Finally, we show the results of motif analysis on three real-world knowledge graphs, compared to the baseline of selecting the most frequent graph patterns.

All code and datasets used in this paper are available online.\footnotemark

\footnotetext{\url{https://github.com/MaestroGraph/motive-rdf}}

@@

Network motifs for unlabeled simple graphs were introduced in \cite{milo2002network}. A more comprehensive overview of the related literature can be found in \cite{bloem2017large}. In \cite{bloem2017large}, the principle of Minimum Description Length (MDL) was first connected to motif analysis. However, the idea had earlier been exploited for detecting meaningful subgraphs in the SUBDUE algorithm \cite{cook1994substructure}.

A few other methods have been proposed for inducing the structure of a given knowledge graph in terms of subgraphs. In \cite{pham2015deriving}, the authors use the principle of characteristic sets to characterize a knowledge graph in terms of the star patterns it contains. In \cite{pham2016exploiting}, they show that the majority of the LOD cloud can be efficiently described using such principles, showing the highly tabular structure of many knowledge graphs. In \cite{volker2011statistical}, association rule mining is used to induce basic patterns in the graph.

@@

\label{section:preliminaries}

\paragraph{Minimum Description Length}
Our method is based on the MDL principle: we should favour models that compress the data. We will show briefly how this intuition can be made mathematically precise. For more details, we refer the reader to \cite{grunwald2007minimum} for MDL in general, and to \cite{bloem2018tutorial}, for a more extensive discussion these principles in the domain of graph analysis.

Let $\B$ be the set of all finite-length binary strings. We use $|b|$ to represent the length of $b \in \B$. Let $\log(x) = \log_2(x)$. A \emph{code} for a set of objects $\cal X$  is an injective function $f: {\cal X} \to \B$. All codes in this paper are \emph{prefix-free}: no code word is the prefix of another. We will denote a \emph{codelength function} with the letter $L$, ie. $L(x) = |f(x)|$. We commonly compute $L(x)$ directly, without first computing $f(x)$.

There is a strong relation between codes and probability distributions: for each probability distribution $p$ on $\cal X$, there exists a prefix-free code $L$ such that for all $x \in \cal X$: $- \log p(x) \leq L(x) < -\log p(x) + 1$. Inversely, for every prefix-free code $L$ for $\cal X$, there exists a probability distribution $p$ such that for all $x \in \cal X$: $p(x) = 2^{-L(x)}$. For proofs, see \cite[Section~3.2.1]{grunwald2007minimum} or \cite[Theorem~5.2.1]{cover2006elements}.

% To explain the intuition, note that we can easily transform a code $L$ into a sampling algorithm for $p$ by feeding the decoding function random bits until it produces an output. To transform a probability distribution to a code, techniques like arithmetic coding \cite{rissanen1979arithmetic} can be used.\footnotemark 
%\footnotetext{
%As explained in \cite[page 96]{grunwald2007minimum}, the fact that $-\log p^*(x)$ is real-valued and $L^*(x)$ is integer-valued can be safely ignored and we may \emph{identify} codes with probability distributions, allowing codes to take non-integer values. 
%}

@@

\paragraph{Relevance testing} We will use the MDL principle to perform a hypothesis test. Assume we have some data $x \in \B$ and a null hypothesis stating that it was sampled from distribution $p^\text{null}$ (with corresponding code $L^\text{null}$). A simple but crucial result, known as the \emph{no-hypercompression inequality} \cite[p103]{grunwald2007minimum} tells us that the probability of sampling any data $x$ from $p^\text{null}$ that can be described in less than $L^\text{null}(x) - k$ or more bits, \emph{using any code} is less than $2^{-k}$.
Thus, we can reject the hypothesis that the data was sampled from $p^\text{null}$ by designing a code which compresses the data better than $L^\text{null}$ by, say, 10 bits and rejecting the null hypothesis with confidence $2^{-10}$. For a longer, more intuitive explanation of this principle in pattern induction, we refer the reader to \cite{bloem2018tutorial}.

Note that when we use this procedure to find motifs, we are not providing statistical evidence for the hypothesis that the motif is ``correct'' \cite[Section~4.1]{bloem2018tutorial}. We are simply using the principle of hypothesis testing as a \emph{heuristic} for pattern mining. The only assertion we are proving (in a statistical sense) is that the data did not come from the null model.

\paragraph{Common codes}
In the construction of our graph codes, we require some simpler codes as building blocks. First, when we store any positive integer $n$, we do so with the code corresponding to the distribution $p^\N(n) = 1/(n(n+1))$, and denote it $L^\N(n)$. For nonnegative numbers we add 1 to the argument. For the full range of integers ($L^\Z$), we add an extra bit for the sign, and then use the first code for negative integers and the second for positive ones.

We will often need to encode \emph{sequences} of integers as well. These will be highly skewed, with only a subset of integers occurring frequently, and others occurring infrequently or not at all. As noted in \cite{de2016names} a code based on the Pitman-Yor model \cite{pitman1997two} is very effective in such situations. Let $S = \langle S_1, ..., S_n\rangle$ be a sequence of integers of length $n$. We first store its members $m(S)$ in the order in which they occur: we first store $n$ and the first member using $L^\N$ and then store each subsequent member by encoding the distance to the previous member using $L^\Z$. Having encoded the members of $S$ we can store the sequence itself using the Pitman-Yor model as follows.

Let $f(A, B)$ be the frequency of symbol $A$ in sequence $B$. We then store the complete sequence using the code corresponding to the following distribution:
\[
p(S) = \prod_{i \in [1,k]} p(S_i\mid S_{1:i-1})\;\;\text{with}\;\;
p(S_i \mid S') = 
\begin{cases}
\frac{\alpha - d |m(S')|}{|m(S')| + \alpha} & \text{if} f(S_i, S') = 0 \\ 
\frac{f(S_i, S') - d}{|m(S')| + \alpha} & \text{otherwise}
\end{cases}
\]
See \cite{de2016names} for a more intuitive explanation. In all experiments we use $\alpha = 0.5$, $d=0.1$. We will refer to the total resulting codelength as $L^{PY}(S)$.

\section{Method}

We will first give a precise definition of a knowledge graph as used in this paper. We will then describe the null model which is used both as a point of comparison in our hypothesis test, and within the motif code to compress the remainder of the graph. Next, we describe how to compress a graph using a given motif, and a set of instances. Finally, we will describe how to search for likely motifs using simulated annealing.
%
%We analyse the \emph{structure} of knowledge graphs only, ignoring any meaning in relation to other graphs, encoded in the content of names or literals, or from ontology languages. 
Specifically, we model a knowledge graph as a multigraph with nodes and edges labeled with integers that map to entities and relations. This mapping is stored, but only the integer-labeled graph is modelled.\footnotemark
\footnotetext{For practitioners this restriction is not noticeable, as the indices can simply be mapped back to the original strings when the found motifs are presented.}

A \emph{knowledge graph} $G$, is a tuple $G = (v_G, r_G, E_G)$. $v_G \in \N$ is the number of nodes in the graph, and $r_G \in \N$ is the number of relations. We define the nodeset of $G$ as $V_G = \{0, \ldots, v_G-1\}$ and the relation-set as $R_G = \{0, \ldots, r_G\}$. The \emph{tripleset} $E_G \subset V_G \times R \times V_G$ determines the edges of the graph and their labels: each triple $(s, r, o) \in E_G$ encodes the originated node $s$, the target node $o$ and the relation $r$ of an edge in the graph.

This definition is compatible with RDF data. We interpret literals as nodes, considered the same node if they are expressed by the same string.

%As shown in Figure~\ref{figure:example}, the patterns that we aim to find are partially labeled: some nodes and edges are labeled as \emph{variables}. We will represent these with negative integers.

A \emph{pattern} $M$ for graph $G$ is a tuple $(V_M, R_M, G, E_M)$. Let $v_M$ and $r_M$ indicate the number of variable nodes and variable links in $M$ respectively, then $V_M \subseteq \{-v'_M, \ldots, v_G-1\}$ and $R_M \subseteq \{-(r'_M+v'M), \ldots,-v'M, 0,\ldots, r_G-1\}$, with $E_M \subset V_M \times R_M \times V_M$ representing the edges as before. That is; nodes can be labeled either with nonnegative integers referring to $G$'s nodes or with negative integers representing a variable node, and similar for relations. The negative integers are always contiguous within a single pattern, with the highest representing the node labels and the lowest representing the edge labels

An \emph{instance} for $M$ in $G$ is a pair of sequences of integers: $I = (I^n, I^r)$. $I^n$ is a sequence of distinct integers of length $v'_M$. $I^r$ is a sequence of non-distinct integers of length $r'_M$. For each edge $(s, p, o) \in E_M$ with any or all of $s$, $p$ and $o$ negative, there is a corresponding link in $E_G$ with a negative $s$ replaced by $I^n_{-s}$, a negative $o$ replaced by $I^n_{-o}$, and a negative $p$ replaced by $I^r_{-p - v'_M}$. Put simply: for a pattern to match, variable edges marked with the same negative integer, must map to the same relation in order for the pattern to match, but variable links labeled with different negative integers \emph{may} map to the same relation. Variable nodes are always labeled distinctly and may never map to the same node in $G$. An instance describes a subgraph of $G$ that \emph{matches} the pattern $M$. Each edge in the motif may only match one edge in the graph. In other words, the occurrence of the motif in the graph must have as many edges as the motif itself.\footnotemark

\footnotetext{In this aspect our definition differs from the SPARQL Basic Graph Pattern. Patterns for which this distinction is relevant are rare, and patterns returned by our method are still compatible with SPARQL.}

\subsection{Null model}

For a proper hypothesis test, we must compare the compression achieved by our motif code to the compression under a general model for knowledge graphs: a null model.

The most common null model in classical motif analysis is the degree-sequence model (also known as the configuration model \cite{newman2010networks}): a uniform distribution over all graphs with a particular degree sequence. We extend this to knowledge graphs by also including the degree of each relation: that is, degree of a relation is the frequency with which it occurs in the tripleset. Let a \emph{degree sequence} $D$ of length $n$ be a triple of three integer sequences: $(D^\text{in}, D^\text{rel}, D^\text{out})$. If $D$ is the degree sequence of a graph, then node $i$ has $D^\text{in}_i$ incoming links,  $D^\text{out}_i$ outgoing links and for each relation $r$, there are $D^\text{rel}_r$ triples.

Let $\G_D$ be the set of all graphs with degree sequence $D$. Then the degree-sequence model can be expressed simply as
\[
p^\text{DS}(G) = \frac{1}{|\G_D|}
\]
for any $G$ that satisfies $D$ and $p(G) = 0$ otherwise. Unfortunately, there is no efficient way to compute $|\G_D|$ and even approximations tend to be costly for large graphs. Following the approach in \cite{bloem2017large}, we define a fast approximation to the configuration model, which works well in practice for motif detection.

@@

We can thus use 
\[
p^\text{EL}_D(G) =  \frac{m!}{{m \choose {D_1^\text{out}, \ldots, D_n^\text{out}} }
 {m \choose {D_1^\text{rel}, \ldots, D_{|R_G|}^\text{rel}} }
 {m \choose {D_1^\text{in}, \ldots, D_n^\text{in}} }} \leq p^\text{DS}(G) \text{.}
\]

Filling in the definition of the multinomial coefficient, and rewriting, we get a codelength of:
\begin{align*}
- \log p^\text{EL}_D(G) = 2 \log(m!)
- \sum_i \log(D_i^\text{in}!)- \sum_i \log(D_i^\text{rel}!)- \sum_i \log(D_i^\text{out}!)
\end{align*}

as an approximation for the DS model. We call this the edgelist (EL) model. It gives a probability that always lower-bounds the configuration model, since it affords some probability mass to graphs that cannot exist. Experiments in the classical motif setting have shown that the EL model is an acceptable proxy for the DS model \cite{bloem2017large}, especially considering the extra scalability it affords.

%\footnotetext{Note that we cannot simply think of $p^\text{EL}$ as a uniform model for graphs containing multiple links, since we can only divide by $m!$ if we know that all triples are unique.}
% 
 \paragraph{Encoding D} In order to encode a graph with $L^\text{EL}_D$, we must first encode $D$.\footnotemark~For each of the three sequences $D'$ in $D$ we use the following model:
\[
 p(D') = \prod_i q^{\N}(D'_i)\;\;\;\;\;
 L(D') = - \sum_i \log q^{\N}(D'_i)
\]

\footnotetext{Or, equivalently, to make $p^\text{EL}$ a complete distribution on all graphs, we must provide it with a prior on $D$.}
